\textbf{SPECFEM3D}\\ \textbf{User Manual}

### 2010-11-16 doc/USER_MANUAL/manual_SPECFEM3D.tex
added a note about the cover page, which will need to be updated
modifies SPECFEM3D attenuation to precompute needed arrays in xgenerate_databases; minor updates in manual

@@

\begin{itemize}
\item the simulation type control parameters: \texttt{SIMULATION\_TYPE}
and \texttt{SAVE\_FORWARD}
\item the time step parameters \texttt{NSTEP} and \texttt{DT}
\item the attenuation control parameters \texttt{ATTENUATION} and \texttt{USE\_OLSEN\_ATTENUATION}
\item the absorbing boundary control parameter \texttt{ABSORBING\_CONDITIONS}
\item the movie control parameters \texttt{MOVIE\_SURFACE}, \texttt{MOVIE\_VOLUME},
and \texttt{NTSTEPS\_BETWEEN\_FRAMES}
\item the ShakeMap\textregistered option \texttt{CREATE\_SHAKEMAP}
\item the output information parameters \texttt{NTSTEP\_BETWEEN\_OUTPUT\_INFO}
and \texttt{NTSTEP\_BETWEEN\_OUTPUT\_}~\\
\texttt{SEISMOS}
\item the \texttt{PRINT\_SOURCE\_TIME\_FUNCTION} flags
\end{itemize}
Any other change to the \texttt{Par\_file} implies rerunning both
the database generator \texttt{xgenerate\_databases} and the solver \texttt{xspecfem3D}.

@@

\begin{enumerate}
\item First, run a regular forward simlation (\texttt{SIMULATION\_TYPE =
1} and \texttt{SAVE\_FORWARD = .false.}). You can automatically set
these two variables using the \texttt{\small UTILS/change\_simulation\_type.pl}
script:

\begin{lyxcode}
UTILS/change\_simulation\_type.pl~-f~
\end{lyxcode}
and then collect the recorded seismograms at all the stations given
in \texttt{in\_data\_files/STATIONS}.

@@

\begin{itemize}
\item Depending on what type of misfit function is used for the source inversion,
adjoint sources need to be computed from the original recorded seismograms
for the selected stations and saved in the \texttt{SEM/} directory
with the format \texttt{STA.NT.BH?.adj}, where \texttt{STA}, \texttt{NT}
are the station name and network code given in the \texttt{in\_data\_files/STATIONS\_ADJOINT}
file, and \texttt{BH?} represents the component name of a particular
adjoint seismogram.
\item The adjoint seismograms are in the same format as the original seismogram
(\texttt{STA.NT.BH?.sem?}), with the same start time, time interval
and record length.
\end{itemize}
\item Notice that even if you choose to time reverse only one component
from one specific station, you still need to supply all three components
because the code is expecting them (you can set the other two components
to be zero).
\item Also note that since time-reversal is done in the code itself, no
explicit time-reversing is needed for the preparation of the adjoint
sources, i.e., the adjoint sources are in the same forward time sense
as the original recorded seismograms.
\end{enumerate}
\item \textbf{Set the related parameters and run the adjoint simulation}\\
In the \texttt{in\_data\_files/Par\_file}, set the two related parameters to
be \texttt{SIMULATION\_TYPE = 2} and \texttt{SAVE\_FORWARD = .false.}.
More conveniently, use the scripts \texttt{UTILS/change\_simulation\_type.pl}
to modify the \texttt{Par\_file} automatically (\texttt{change\_simulation\_type.pl
-a}). Then run the solver to launch the adjoint simulation.
\item \textbf{Collect the seismograms at the original source location}

@@

\begin{itemize}
\item In the case of travel-time finite-frequency kernel for one source-receiver
pair, i.e., point source from the \texttt{CMTSOLUTION}, and one station
in the \texttt{STATIONS\_ADJOINT} list, we supply a sample program
in \texttt{UTILS/xcut\_velocity} to cut a certain portion of the original
displacement seismograms and convert it into the proper adjoint source
to compute the finite-frequency kernel.

@@

With the successful forward simulation and the adjoint source ready
in \texttt{SEM/}, set \texttt{SIMULATION\_TYPE = 3} and \texttt{SAVE\_FORWARD
= .false.} in the \texttt{Par\_file(change\_simulation\_type.pl -b),}
and rerun the solver.

@@

A Perl script \texttt{UTILS/slice\_number.pl} that calls MATLAB can
help to figure out the slice numbers that lie along the great circle
path (both the minor and major arcs), as well as the slice numbers
required to produce a full picture of the inner core if your kernel
also illuminates the inner core.

\begin{enumerate}
\item On machines where you have MATLAB access, copy the \texttt{CMTSOLUTION}
file, \texttt{STATIONS\_ADJOINT}, and \texttt{Par\_file}, and run:

\begin{lyxcode}
{\small UTILS/slice\_number.pl~Par\_file~output\_solver.txt~slice\_file}{\small \par}
\end{lyxcode}
which will generate a \texttt{slices\_file}.

@@

\begin{enumerate}
\item You can use or modify the script \texttt{UTILS/copy\_databases.pl}
to accomplish this:

\begin{lyxcode}
{\small UTILS/copy\_database.pl~slice\_file~lsf\_machine\_file~filename~{[}jobid]}{\small \par}
\end{lyxcode}
where \texttt{\small lsf\_machine\_file} is the machine file generated
by the LSF scheduler, \texttt{\small filename} is the kernel name
(e.g., \texttt{\small rho\_kernel}, \texttt{\small alpha\_kernel}
and \texttt{\small beta\_kernel}), and the optional \texttt{\small jobid}
is the name of the subdirectory under \texttt{\small LOCAL\_PATH}
where all the kernel files are stored.

@@

We use an auxilliary program \texttt{combine\_paraview\_data.f90}
to combine the kernel files from all slices into one mesh file.

\begin{enumerate}
\item Compile it in the global code directory:

\begin{lyxcode}
{\footnotesize make~combine\_paraview\_data~}{\footnotesize \par}

{\footnotesize xcombine\_paraview\_data~slice\_list~filename~input\_dir~output\_dir~high/low-resolution~}{\footnotesize \par}
\end{lyxcode}
where \texttt{input\_dir} is the directory where all the individual
kernel files are stored, and \texttt{output\_dir} is where the mesh
file will be written.

@@

\begin{enumerate}
\item We next convert the \texttt{.mesh} file into the VTU (Unstructured
grid file) format which can be viewed in ParaView, for example:

\begin{lyxcode}
UTILS/mesh2vtu.pl~-i~file.mesh~-o~file.vtu
\end{lyxcode}
\item Notice that this Perl script uses a program \texttt{mesh2vtu} in the
\texttt{UTILS/mesh2vtu} directory, which further uses the VTK \url{http://www.vtk.org/}
run-time library for its execution. Therefore, make sure you have
them properly set in the script according to your system.
\end{enumerate}
\item \textbf{Copy over the source and receiver .vtk file}

@@

\begin{itemize}
\item The processors/nodes to be used for each run are assigned dynamically
by the scheduler, based on availability. Therefore, in order for the
\texttt{xgenerate\_databases} and the \texttt{xspecfem3D} executables (or between successive runs of the solver) to
have access to the same database files (if they are stored on hard
drives local to the nodes on which the code is run), they must be
launched in sequence as a single job.
\item On some systems, the nodes to which running jobs are assigned are
not configured for compilation. It may therefore be necessary to pre-compile
both the \texttt{xgenerate\_databases} and the \texttt{xspecfem3D} executables.
%obsolete
%A small program provided in the distribution
%called \texttt{\small create\_header\_file.f90} can be used to directly
%create \texttt{\small in\_out\_files/OUTPUT\_FILES/values\_from\_mesher.h} using
%the information in the \texttt{\small in\_data\_files/Par\_file} without having
%to run the mesher (type \texttt{\small `make}{\small{} }\texttt{\small create\_header\_}~\\
%\texttt{\small file}' to compile it and `\texttt{\small xcreate\_header\_file}'
%to run it; refer to the sample scripts below). The solver can now
%be compiled as explained above.
\item One feature of schedulers/queuing systems is that they allow submission
of multiple jobs in a ``launch and forget'' mode. In order to take
advantage of this property, care needs to be taken that output and
intermediate files from separate jobs do not overwrite each other,
or otherwise interfere with other running jobs.
\end{itemize}
Examples of job scripts can be found in \texttt{\small UTILS/Cluster/}
directory and can straightforwardly be modified and adapted to meet
more specific running needs.

We describe here in some detail a job submission procedure for the
Caltech 1024-node cluster, CITerra, under the LSF scheduling system.
We consider the submission of a regular forward simulation using the internal mesher
to create mesh partitions. The two
main scripts are \texttt{\small run\_lsf.bash}, which compiles the
Fortran code and submits the job to the scheduler, and \texttt{\small go\_mesher\_solver\_lsf\_basin.forward},
which contains the instructions that make up
the job itself. These scripts can be found in \texttt{\small UTILS/Cluster/lsf/} directory

@@

\begin{enumerate}
\item First the script directs the scheduler to save its own output and
output from \texttt{stdout} into ~\\
\texttt{\small in\_out\_files/OUTPUT\_FILES/\%J.o},
where \texttt{\%J} is short-hand for the job-ID; it also tells the
scheduler what version of \texttt{mpich} to use (\texttt{mpich\_gm})
and how to name this job (\texttt{go\_mesher\_solver\_lsf}).
\item The script then creates a list of the nodes allocated to this job
by echoing the value of a dynamically set environment variable \texttt{LSB\_MCPU\_HOSTS}
and parsing the output into a one-column list using the Perl script
\texttt{UTILS/remap\_lsf\_machines.pl}. It then creates a set of scratch
directories on these nodes (\texttt{\small /scratch/}~\\
\texttt{\small \$USER/DATABASES\_MPI}) to be used as the \texttt{LOCAL\_PATH}
for temporary storage of the database files. The scratch directories
are created using \texttt{shmux}, a shell multiplexor that can execute
the same commands on many hosts in parallel. \texttt{shmux} is available
from Shmux \url{web.taranis.org/shmux/}. Make sure that the \texttt{LOCAL\_PATH}
parameter in \texttt{in\_data\_files/Par\_file} is also set properly.
\item The next portion of the script launches the mesher, database generator and then the solver
using \texttt{run\_lsf.bash}.
\item The final portion of the script collects the seismograms and performs
clean up on the nodes, using the Perl scripts \texttt{collect\_seismo\_lsf\_multi.pl}
and \texttt{cleanmulti.pl}.
\end{enumerate}
\begin{lyxcode}
\#!/bin/bash~-v~

@@

Several post-processing scripts/programs are provided in the \texttt{UTILS/}
directory, and most of them need to be adjusted when used on different
systems, for example, the path of the executable programs. Here we
only list a few of the available scripts and provide a brief description, and
you can either refer to the related sections for detailed usage or,
in a lot of cases, type the script/program name without arguments
for its usage.

@@

In many cases, the SEM synthetics are calculated and compared to data
seismograms recorded at seismic stations. Since the SEM synthetics
are accurate for a certain frequency range, both the original data
and the synthetics need to be processed before a comparison can be
made. We generally use the following scripts provided in the ~\\
\texttt{UTILS/seis\_process/} directory:

@@

You do have two options to change this default output behavior, given in the main constants file
\texttt{constants.h} located in \texttt{src/shared/} directory:
\begin{description}
\item [{\texttt{WRITE\_SEISMOGRAMS\_BY\_MASTER}}] Set to \texttt{.true.} to have only the master process writing out seismograms. This can be useful on a cluster, where only the master process node has access to the output directory.
\item [{\texttt{USE\_OUTPUT\_FILES\_PATH}}] Set to \texttt{.false.} to have the seismograms output to \texttt{LOCAL\_PATH} directory specified in the main parameter file \texttt{in\_data\_files/Par\_file}.
In this case, you could collect the synthetics onto the frontend using the \texttt{collect\_seismo\_lsf\_multi.pl} script
located in the \texttt{UTILS/Cluster/lsf/} directory. The usage of the script would be e.g.:

@@

After all the simulations are done, the seismograms are collected,
and the useful database files are copied to the frontend, you may
need to clean the local scratch disk for the next simulation. This
is especially important in the case of kernel simulation,
where very large files are generated for the absorbing boundaries
to help with the reconstruction of the regular forward wavefield.
A sample script is provided in \texttt{UTILS/}:

>>>>

\toall{We should also update the Adobe Illustrator / PDF cover page as well at some point}

@@

\begin{itemize}
\item the simulation type control parameters: \texttt{SIMULATION\_TYPE}
and \texttt{SAVE\_FORWARD}
\item the time step parameters \texttt{NSTEP} and \texttt{DT}
%\item the attenuation control parameters \texttt{ATTENUATION} and \texttt{USE\_OLSEN\_ATTENUATION}
\item the absorbing boundary control parameter \texttt{ABSORBING\_CONDITIONS}
\item the movie control parameters \texttt{MOVIE\_SURFACE}, \texttt{MOVIE\_VOLUME},
and \texttt{NTSTEPS\_BETWEEN\_FRAMES}
\item the ShakeMap\textregistered option \texttt{CREATE\_SHAKEMAP}
\item the output information parameters \texttt{NTSTEP\_BETWEEN\_OUTPUT\_INFO}
and \texttt{NTSTEP\_BETWEEN\_OUTPUT\_}~\\
\texttt{SEISMOS}
\item the \texttt{PRINT\_SOURCE\_TIME\_FUNCTION} flags
\end{itemize}
Any other change to the \texttt{Par\_file} implies rerunning both
the database generator \texttt{xgenerate\_databases} and the solver \texttt{xspecfem3D}.

@@

\begin{enumerate}
\item First, run a regular forward simlation (\texttt{SIMULATION\_TYPE =
1} and \texttt{SAVE\_FORWARD = .false.}). You can automatically set
these two variables using the \texttt{\small utils/change\_simulation\_type.pl}
script:

\begin{lyxcode}
utils/change\_simulation\_type.pl~-f~
\end{lyxcode}
and then collect the recorded seismograms at all the stations given
in \texttt{in\_data\_files/STATIONS}.

@@

\begin{itemize}
\item Depending on what type of misfit function is used for the source inversion,
adjoint sources need to be computed from the original recorded seismograms
for the selected stations and saved in the \texttt{in\_out\_files/SEM/} directory
with the format \texttt{STA.NT.BH?.adj}, where \texttt{STA}, \texttt{NT}
are the station name and network code given in the \texttt{in\_data\_files/STATIONS\_ADJOINT}
file, and \texttt{BH?} represents the component name of a particular
adjoint seismogram.
\item The adjoint seismograms are in the same format as the original seismogram
(\texttt{STA.NT.BH?.sem?}), with the same start time, time interval
and record length.
\end{itemize}
\item Notice that even if you choose to time reverse only one component
from one specific station, you still need to supply all three components
because the code is expecting them (you can set the other two components
to be zero).
\item Also note that since time-reversal is done in the code itself, no
explicit time-reversing is needed for the preparation of the adjoint
sources, i.e., the adjoint sources are in the same forward time sense
as the original recorded seismograms.
\end{enumerate}
\item \textbf{Set the related parameters and run the adjoint simulation}\\
In the \texttt{in\_data\_files/Par\_file}, set the two related parameters to
be \texttt{SIMULATION\_TYPE = 2} and \texttt{SAVE\_FORWARD = .false.}.
More conveniently, use the scripts \texttt{utils/change\_simulation\_type.pl}
to modify the \texttt{Par\_file} automatically (\texttt{change\_simulation\_type.pl
-a}). Then run the solver to launch the adjoint simulation.
\item \textbf{Collect the seismograms at the original source location}

@@

\begin{itemize}
\item In the case of travel-time finite-frequency kernel for one source-receiver
pair, i.e., point source from the \texttt{CMTSOLUTION}, and one station
in the \texttt{STATIONS\_ADJOINT} list, we supply a sample program
in \texttt{utils/cut\_velocity/xcut\_velocity} to cut a certain portion of the original
displacement seismograms and convert it into the proper adjoint source
to compute the finite-frequency kernel.

@@

With the successful forward simulation and the adjoint source ready
in the \texttt{in\_out\_files/SEM/} directory, set \texttt{SIMULATION\_TYPE = 3} and \texttt{SAVE\_FORWARD
= .false.} in the \texttt{Par\_file(change\_simulation\_type.pl -b),}
and rerun the solver.

@@

A Perl script \texttt{slice\_number.pl} located in directory \texttt{utils/Visualization/Paraview/} can
help to figure out the slice numbers that lie along the great circle
path. It applies to meshes created with the internal mesher \texttt{xmeshfem3D}.

\begin{enumerate}
\item On machines where you have access to the script, copy the \texttt{Mesh\_Par\_file}, and
\texttt{output\_solver} files, and run:

\begin{lyxcode}
{\small slice\_number.pl~Mesh\_Par\_file~output\_solver.txt~slice\_file}{\small \par}
\end{lyxcode}
which will generate a \texttt{slices\_file}.

@@

\begin{enumerate}
\item You can use or modify the script \texttt{utils/copy\_basin\_database.pl}
to accomplish this:

\begin{lyxcode}
{\small utils/copy\_database.pl~slice\_file~lsf\_machine\_file~filename~{[}jobid]}{\small \par}
\end{lyxcode}
where \texttt{\small lsf\_machine\_file} is the machine file generated
by the LSF scheduler, \texttt{\small filename} is the kernel name
(e.g., \texttt{\small rho\_kernel}, \texttt{\small alpha\_kernel}
and \texttt{\small beta\_kernel}), and the optional \texttt{\small jobid}
is the name of the subdirectory under \texttt{\small LOCAL\_PATH}
where all the kernel files are stored.

@@

We use an auxilliary program \texttt{combine\_vol\_data.f90}
to combine the kernel files from all slices into one mesh file.

\begin{enumerate}
\item Compile it in the root directory:

\begin{lyxcode}
{\footnotesize make~combine\_vol\_data~}{\footnotesize \par}

{\footnotesize ./bin/xcombine\_vol\_data~slice\_list~filename~input\_dir~output\_dir~high/low-resolution~}{\footnotesize \par}
\end{lyxcode}
where \texttt{input\_dir} is the directory where all the individual
kernel files are stored, and \texttt{output\_dir} is where the mesh
file will be written.

@@

\begin{enumerate}
\item We next convert the \texttt{.mesh} file into the VTU (Unstructured
grid file) format which can be viewed in ParaView. For this task, you can use and modify the script \texttt{mesh2vtu.pl}
located in directory \texttt{utils/Visualization/Paraview/}, for example:

\begin{lyxcode}
mesh2vtu.pl~-i~file.mesh~-o~file.vtu
\end{lyxcode}
\item Notice that this Perl script uses a program \texttt{mesh2vtu} in the
\texttt{utils/Visualization/Paraview/mesh2vtu} directory, which further uses the VTK \url{http://www.vtk.org/}
run-time library for its execution. Therefore, make sure you have
them properly set in the script according to your system.
\end{enumerate}
\item \textbf{Copy over the source and receiver .vtk file}

@@

\begin{itemize}
\item The processors/nodes to be used for each run are assigned dynamically
by the scheduler, based on availability. Therefore, in order for the
\texttt{xgenerate\_databases} and the \texttt{xspecfem3D} executables (or between successive runs of the solver) to
have access to the same database files (if they are stored on hard
drives local to the nodes on which the code is run), they must be
launched in sequence as a single job.
\item On some systems, the nodes to which running jobs are assigned are
not configured for compilation. It may therefore be necessary to pre-compile
both the \texttt{xgenerate\_databases} and the \texttt{xspecfem3D} executables.
%obsolete
%A small program provided in the distribution
%called \texttt{\small create\_header\_file.f90} can be used to directly
%create \texttt{\small in\_out\_files/OUTPUT\_FILES/values\_from\_mesher.h} using
%the information in the \texttt{\small in\_data\_files/Par\_file} without having
%to run the mesher (type \texttt{\small `make}{\small{} }\texttt{\small create\_header\_}~\\
%\texttt{\small file}' to compile it and `\texttt{\small xcreate\_header\_file}'
%to run it; refer to the sample scripts below). The solver can now
%be compiled as explained above.
\item One feature of schedulers/queuing systems is that they allow submission
of multiple jobs in a ``launch and forget'' mode. In order to take
advantage of this property, care needs to be taken that output and
intermediate files from separate jobs do not overwrite each other,
or otherwise interfere with other running jobs.
\end{itemize}
Examples of job scripts can be found in the \texttt{\small utils/Cluster/}
directory and can straightforwardly be modified and adapted to meet
more specific running needs.

We describe here in some detail a job submission procedure for the
Caltech 1024-node cluster, CITerra, under the LSF scheduling system.
We consider the submission of a regular forward simulation using the internal mesher
to create mesh partitions. The two
main scripts are \texttt{\small run\_lsf.bash}, which compiles the
Fortran code and submits the job to the scheduler, and \texttt{\small go\_mesher\_solver\_lsf\_basin.forward},
which contains the instructions that make up
the job itself. These scripts can be found in \texttt{\small utils/Cluster/lsf/} directory

@@

\begin{enumerate}
\item First the script directs the scheduler to save its own output and
output from \texttt{stdout} into ~\\
\texttt{\small in\_out\_files/OUTPUT\_FILES/\%J.o},
where \texttt{\%J} is short-hand for the job-ID; it also tells the
scheduler what version of \texttt{mpich} to use (\texttt{mpich\_gm})
and how to name this job (\texttt{go\_mesher\_solver\_lsf}).
\item The script then creates a list of the nodes allocated to this job
by echoing the value of a dynamically set environment variable \texttt{LSB\_MCPU\_HOSTS}
and parsing the output into a one-column list using the Perl script
\texttt{utils/Cluster/lsf/remap\_lsf\_machines.pl}. It then creates a set of scratch
directories on these nodes (\texttt{\small /scratch/}~\\
\texttt{\small \$USER/DATABASES\_MPI}) to be used as the \texttt{LOCAL\_PATH}
for temporary storage of the database files. The scratch directories
are created using \texttt{shmux}, a shell multiplexor that can execute
the same commands on many hosts in parallel. \texttt{shmux} is available
from Shmux \url{web.taranis.org/shmux/}. Make sure that the \texttt{LOCAL\_PATH}
parameter in \texttt{in\_data\_files/Par\_file} is also set properly.
\item The next portion of the script launches the mesher, database generator and then the solver
using \texttt{run\_lsf.bash}.
\item The final portion of the script collects the seismograms and performs
clean up on the nodes, using the Perl scripts \texttt{collect\_seismo\_lsf\_multi.pl}
and \texttt{cleanmulti.pl}.
\end{enumerate}
\begin{lyxcode}
\#!/bin/bash~-v~

@@

Several post-processing scripts/programs are provided in the \texttt{utils/}
directory, and most of them need to be adjusted when used on different
systems, for example, the path of the executable programs. Here we
only list a few of the available scripts and provide a brief description, and
you can either refer to the related sections for detailed usage or,
in a lot of cases, type the script/program name without arguments
for its usage.

@@

In many cases, the SEM synthetics are calculated and compared to data
seismograms recorded at seismic stations. Since the SEM synthetics
are accurate for a certain frequency range, both the original data
and the synthetics need to be processed before a comparison can be
made. We generally use the following scripts provided in the ~\\
\texttt{utils/seis\_process/} directory:

@@

You do have two options to change this default output behavior, given in the main constants file
\texttt{constants.h} located in \texttt{src/shared/} directory:
\begin{description}
\item [{\texttt{WRITE\_SEISMOGRAMS\_BY\_MASTER}}] Set to \texttt{.true.} to have only the master process writing out seismograms. This can be useful on a cluster, where only the master process node has access to the output directory.
\item [{\texttt{USE\_OUTPUT\_FILES\_PATH}}] Set to \texttt{.false.} to have the seismograms output to \texttt{LOCAL\_PATH} directory specified in the main parameter file \texttt{in\_data\_files/Par\_file}.
In this case, you could collect the synthetics onto the frontend using the \texttt{collect\_seismo\_lsf\_multi.pl} script
located in the \texttt{utils/Cluster/lsf/} directory. The usage of the script would be e.g.:

@@

After all the simulations are done, the seismograms are collected,
and the useful database files are copied to the frontend, you may
need to clean the local scratch disk for the next simulation. This
is especially important in the case of kernel simulation,
where very large files are generated for the absorbing boundaries
to help with the reconstruction of the regular forward wavefield.
A sample script is provided in \texttt{utils/}:
